### 2021-11-16 thesis/tex/oscillations.tex
update

Experimental measurements show that $\theta_{13}$ is the smallest of the mixing angles~\cite{nufit}, causing 
$\nm$ to primarily mix into $\nt$. The mixing matrix from~\ref{PMNS_def} takes the values
\begin{align}\label{eq:Uvalues}
   U = \begin{pmatrix}
       U_{e 1} & U_{e2} & U_{e3} \\
       U_{\mu 1} & U_{\mu 2} & U_{\mu 3} \\
       U_{\tau 1} & U_{\tau 2} & U_{\tau 3}
   \end{pmatrix} 
   = \begin{pmatrix}
       0.825 & 0.545 & 0.149 \\
       -0.455 & 0.485 & 0.746 \\
       0.334 & -0.684 & 0.649
   \end{pmatrix} \,,
\end{align}
which use to we plot the $P_{\mu \beta}$ probabilities in Fig.~\ref{fig:vac_osc}. Here, we show the energy spectra of these oscillations in the low \si{\GeV} range for neutrinos that travel \SI{12000}{\km}\footnote{This baseline
is approximately the Earth diameter, and will be the maximum travel distance which we will use later on.}.
In the left-most panel, we see a supressed $\nm \to \ne$ transition. We can understand this behavior by studying Eq.~\ref{eq:Pab} and the 
numerical values of the mixing matrix in Eq.~\ref{eq:Uvalues}.
The smallness of $U_{e3}$ reduces the effect of $\dm[21]$ mass-squared difference,
allowing $\dm[31]$ to drive the oscillations. 
The most important mixing matrix 
elements for $P_{\a\b}$ are $U_{\a3}, U_{\a1}$ and $U_{\b3}, U_{\b1}$. %TODO: write whole 4 product here
%instead of U,U and U,U

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=0.9\textwidth]{figures/w-boson.pdf} 
        \caption{Charged current weak interaction between an electron neutrino and an electron,
        mediated by either a $W^+$ or $W^-$ boson.}
    \end{subfigure}
    \quad
    \begin{subfigure}{0.3\textwidth}
        \vspace{1em}
        \includegraphics[width=1\textwidth]{figures/z-boson.pdf} 
        \caption{Neutral current weak interaction between a neutrino of any flavor and an electron, proton, or neutron,
        mediated by a neutral $Z$ boson.}
    \end{subfigure}
    \caption{Feymann diagrams showing neutrinos participating in the two types of weak interactions according to the Standard Model.}\label{fig:w_and_z}
\end{figure}
So far we have discussed neutrinos travelling in vacuum. Now we will concern ourselves with neutrino oscillations in matter.
The possible interactions are shown in Fig.~\ref{fig:w_and_z}. The left panel shows that the only flavor that can undergo charged current (CC) 
interactions are the electron flavor. This is because the CC interaction requires a neutrino with flavor $\alpha$ to interact with its charged lepton 
partner $\alpha$. Since the Earth doesn't consist of any muons or tau particles, the $\nu_\mu$ and $\nu_\tau$ have no charged leptons to undergo CC interactions with. 
The right panel shows any neutrino flavor interaction via the neutral current (NC) with matter, mediated by the neutral $Z$ boson. The Earth is entirely composed
of electrons, protons, and neutrons. Thus, the fundamental particles composing Earth are electrons, and up and down quarks. These are the 
fermions relevant for our matter oscillations. 
The interaction mediated by the $W$ boson will give rise to an effective matter potential $V_{CC}$, while
the $Z$ boson is responsible for $V_{NC}$.

@@

            &= \sqrt{2}G_F N_e \bar{\nu}_{eE} \gamma^0 \nu_{eL}\,,

@@

where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand compontent of the neutrino fields interact. Thus,
\begin{align}\label{eq:V_CC1}
    V_{CC} = \sqrt{2}G_F N_e\,, \quad H_{CC}\ket{\nu_k} = V_{CC}\ket{\nu_k}\,.
Here we see a crucial difference between the eigenvectors between the vacuum Hamiltonian defined in Eq.~\ref{eq:osc_1} and $H_{CC}$, namely
that the CC interactions happen in the flavor basis rather than in the mass basis\footnote{A similar calculation for the NC reveals the same fact:
neutrinos interact in the flavor basis.}. In other words,
neutrinos propagate in their mass eigenstates but interact in their flavor eigenstate. The mixing of mass eigenstates during propagation
determines if the flavor eigenstate has oscillated or not. The interactions still conserve lepton number since the interaction Lagrangian does,
but the kinetic Lagrangian will not. Hence, neutrinos violate lepton number convervation during propagation.

@@

\subsection{Matter Oscillations}
Since only $\ne$ undergo CC interactions in Earth-like matter, the $V_{CC}$ potential is zero for all other flavors. 
However, since all flavors undergo NC interactions the total matter potential in the flavor basis is
\begin{align}\label{eq:V_matrix}
    V = \begin{bmatrix}
        V_{CC} + V_{NC} & 0 & 0 \\
        0 & V_{NC} & 0 \\
        0 & 0 & V_{NC} \\
    \end{bmatrix} = V_{CC}\, \delta_{\alpha e}\mathbb{I} + V_{NC}\mathbb{I}\,.
\end{align}
Just as in Eq.~\ref{eq:TDSE}, we start with a Hamiltonian that solves the time-dependent Schrödinger equation. This time, let the Hamiltonian be 
\begin{align}
    H = H_0 + H_{I}\,,
\end{align}
where $H_0$ is the Hamiltonian in vacuum, and $H_{I}$ is our interaction Hamiltonian associated with our matter potentials.
Let the wavefunction that describes the $\nu_\a \to \nu_\b$ transition be denoted as
\begin{align}
    \psi_{\a\b}(t) = \bra{\nu_\b}\ket{\nu_\a (t)}\,,
\end{align}
i.e.~the evolution of the state of a neutrino emitted at $t =0$ with flavor $\alpha$ to flavor $\beta$ at time $t$. Now using Eq.~\ref{eq:osc_1} and Eq.~\ref{eq:V_CC1}, we are ready to see what form our Hamiltonians take.
Let us start with the vaccum Hamiltonian $H_0$, and act on its Schrödinger equation with $\bra{\nu_\b}\,$:
\begin{align}
    i \dv{}{t}\ket{\nu_\a (t)} = H_0\ket{\nu_\a (t)} \implies i \dv{}{t}\psi_{\a\b} = \bra{\nu_\b}H_0\ket{\nu_\a (t)}\,.
\end{align}
Reminding ourselves that the vacuum Hamiltonian $H_0$ has eigenstates in the mass basis, we write the following expression where 
use the relations Eq.~\ref{eq:osc_1} and Eq.~\ref{eq:osc_k} to switch between the flavor and mass basis with the PMNS elements:
\begin{align}
    \bra{\nu_\b} H_0 &= \sum_k U_{\b k} \bra{\nu_k}H_0 \nonumber \\
                     &= \sum_k U_{\b k} E_k \bra{\nu_k} \nonumber \\
                     &= \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \bra{\nu_\eta}\,.
\end{align}
Thus,
\begin{align}
    \bra{\nu_\b}H_0\ket{\nu_\a (t)} &= \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \bra{\nu_\eta}\ket{\nu_\a (t)} \nonumber \\
                                    &= \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \psi_{\a\eta}(t)\,.
\end{align}
Using the ultrarelativistic approximation from Eq.~\ref{eq:ultra_rel}:
\begin{align}
    \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \psi_{\a\eta}(t) &= \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_k}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x) \nonumber \\
    &= \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_k}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x)\,.
\end{align}
Use the fact that $\sum_k m_k^2 =  \sum m_1^2 +_k m^2_k - m^2_1 =\sum_k m_1^2 +  \dm_{k1}$ to pull out common terms out of the summation:
\begin{align}\label{eq:t1}
    \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_k}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x) &= \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_1}{2E} + \frac{\dm_{k1}}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x) \nonumber \\
    &= \sum_\eta\sum_k \left(p + \frac{m^2_1}{2E}\right) U_{\b k} U^*_{\eta k} \psi_{\a\eta}(x) \nonumber \\
    &+ \sum_\eta\sum_k U_{\b k} \frac{\dm_{k1}}{2E} U^*_{\eta k} \psi_{\a\eta}(x)\,.
\end{align}
Unitarity gives $ \sum_k U_{\b k} U^*_{\eta k} = \delta_{\eta\b}$, and the first term in the last step of Eq.~\ref{eq:t1} becomes
\begin{align}
    \sum_\eta \left(p + \frac{m^2_1}{2E}\right) \delta_{\beta \eta} \psi_{\a\eta}(x)
    =& \left(p + \frac{m^2_1}{2E}\right) \psi_{\a\b}(x)\,.
\end{align}

@@

\subsection{Two neutrino mixing}
By working in the limit where one of the neutrino masses can be neglected, 
we can reduce the complexity of the analytical solution of neutrino oscillation. 
We now only consider the mixing between two neutrino flavors, say between $\ne, \nm$, and $\nu_1, \nu_2$. In other words, we let $\dm[31] = 0$.

The three-neutrino oscillation Hamiltonian from Eq.~\eqref{eq:H_3gen} %TODO: derive this
now reduces to 
\begin{align}
    H_F &= \frac{1}{2E}(U M^2 U^\dagger + A) \nonumber \\
        &= \frac{1}{2E}\left[U \begin{pmatrix}
            0 & 0 & 0 \\
            0 & \dm_{21} & 0 \\
            0 & 0 & 0
        \end{pmatrix} U^\dagger\right] + \sqrt{2}G_F N_e \begin{pmatrix}
            1 & 0 & 0 \\
            0 & 0 & 0 \\
            0 & 0 & 0
        \end{pmatrix}\,. 
\end{align}
The rotation matrices $R_{ij}$ that constitute $U$ according to Eq.~\eqref{PMNS_def} can be simplified. The rotation $R_{13}$ will not affect our mass matrix, since the former acts in the $1-3$ plane, and the latter sits in the $1-2$ plane. Thus, the reduced mass matrix $\mathbb{M}_{12}$ commutes with $R_{13}$. Analogously, the  rotation matrix $R_{23}$ also commutes with $\mathbb{M}_{12}$.
Moreover, since $SO(3)$ is abelian, its generators commute, i.e.
\begin{align}
    \commutator{R_{ij}}{R_{kl}} = R_{ij}R_{kl} - R_{kl}R_{ij} = 0\,.
\end{align}
Thus,
\begin{align}
    U \mathbb{M}_{12} U^\dagger &=
    R_{23}R_{13}R_{12} \mathbb{M}_{12}  R_{12}^\dagger R_{13}^\dagger  R_{23}^\dagger \nonumber \\
    &= R_{12}R_{23}R_{13} \mathbb{M}_{12}  R_{13}^\dagger R_{23}^\dagger R_{12}^\dagger   \nonumber \\
    &= R_{12}\mathbb{M}_{12} R_{23} R_{13}   R_{13}^\dagger R_{23}^\dagger R_{12}^\dagger  \,.
\end{align}
And due to the fact that the rotation matrices are unitary,
\begin{align}
    R\,R^\dagger = \mathbb{I}\,,
\end{align}
we have that 
\begin{align}
    U \mathbb{M}_{12} U^\dagger &= R_{12}\mathbb{M}_{12}R_{23} R_{13}   R_{13}^\dagger R_{12}^\dagger  R_{23}^\dagger \nonumber \\
    &= R_{12}\mathbb{M}_{12}R_{23} R_{23}^\dagger  R_{12}^\dagger \nonumber \\
    &= R_{12}\mathbb{M}_{12}  R_{12}^\dagger\,.
\end{align}
We have now completely removed the dependence of the mixing angle between the $\nu_1$ and $\nu_3$ mass eigenstates, $\theta_{13}$, and the dependece of the mixing angle between the $\nu_2$ and $\nu_3$ mass eigenstates, $\theta_{23}$.
We can see this clearly by writing our reduced mixing matrix, which is only constituted by the rotation in the $1-2$ plane, $R_{12}$.

\begin{align}
    U &= R_{12}
\begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1 \end{pmatrix} 
\end{align}

@@

Now we need to incorporate the \emph{zenith angle}, here defined as the angle between the neutrino direction of travel and south.
This way, neutrinos traveling through the entire diameter of the Earth up through the South Pole are defined as `up-going', while
neutrinos that start at the South Pole are `down-going'. We will mostly work with the quantity $\cos{(\theta_z)}$.
Since we are interested in the matter effects, some studies are only looking at up-going neutrinos, i.e. neutrinos with
zenith angle $-1 \le \cos{(\theta_z)} \le 0$. However, while down-going neutrinos don't experience matter oscillations (and hardly even vacuum
oscillations due to the short baseline), they can be included to reduce systematic errors.
We now supplement our probability grid from Fig.~\ref{fig:oscillations} with the zenith dimension,
allowing us to fully see the Earth matter effect on the oscillations in full. This is shown in Fig.~\ref{fig:oscillograms}.

>>>>

Experimental measurements show that $\theta_{13}$ is the smallest of the mixing angles~\cite{nufit}, causing 
$\nm$ to primarily mix into $\nt$. The mixing matrix from~\ref{PMNS_def} takes the values
\begin{align}\label{eq:Uvalues}
   U = \begin{pmatrix}
       U_{e 1} & U_{e2} & U_{e3} \\
       U_{\mu 1} & U_{\mu 2} & U_{\mu 3} \\
       U_{\tau 1} & U_{\tau 2} & U_{\tau 3}
   \end{pmatrix} 
   = \begin{pmatrix}
       0.825 & 0.545 & 0.149 \\
       -0.455 & 0.485 & 0.746 \\
       0.334 & -0.684 & 0.649
   \end{pmatrix} \,,
\end{align}
which use to we plot the $P_{\mu \beta}$ probabilities in Fig.~\ref{fig:vac_osc}. Here, we show the energy spectra of these oscillations in the low \si{\GeV} range for neutrinos that travel \SI{12000}{\km}\footnote{This baseline
is approximately the Earth diameter, and will be the maximum travel distance which we will use later on.}.
In the left-most panel, we see a suppressed $\nm \to \ne$ transition. We can understand this behavior by studying Eq.~\ref{eq:Pab} and the 
numerical values of the mixing matrix in Eq.~\ref{eq:Uvalues}.
The smallness of $U_{e3}$ reduces the effect of $\dm[21]$ mass-squared difference,
allowing $\dm[31]$ to drive the oscillations. 
The most important mixing matrix 
elements for $P_{\a\b}$ are $U_{\a3}, U_{\a1}$ and $U_{\b3}, U_{\b1}$. %TODO: write whole 4 product here
%instead of U,U and U,U

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=0.9\textwidth]{figures/w-boson.pdf} 
        \caption{Charged current weak interaction between an electron neutrino and an electron,
        mediated by either a $W^+$ or $W^-$ boson.}
    \end{subfigure}
    \quad
    \begin{subfigure}{0.3\textwidth}
        \vspace{1em}
        \includegraphics[width=1\textwidth]{figures/z-boson.pdf} 
        \caption{Neutral current weak interaction between a neutrino of any flavor and an electron, proton, or neutron,
        mediated by a neutral $Z$ boson.}
    \end{subfigure}
    \caption{Feymann diagrams showing neutrinos participating in the two types of weak interactions according to the Standard Model.}\label{fig:w_and_z}
\end{figure}
So far we have discussed neutrinos traveling in vacuum. Now we will concern ourselves with neutrino oscillations in matter.
The possible interactions are shown in Fig.~\ref{fig:w_and_z}. The left panel shows that the only flavor that can undergo charged current (CC) 
interactions are the electron flavor. This is because the CC interaction requires a neutrino with flavor $\alpha$ to interact with its charged lepton
partner $\alpha$. Since the Earth doesn't consist of any muons or tau particles, the $\nu_\mu$ and $\nu_\tau$ have no charged leptons to undergo CC interactions with. 
The right panel shows any neutrino flavor interaction via the neutral current (NC) with matter, mediated by the neutral $Z$ boson. The Earth is entirely composed
of electrons, protons, and neutrons. Thus, the fundamental particles composing Earth are electrons, and up and down quarks. These are the
fermions relevant for our matter oscillations. In these two diagrams, we see that the CC interactions happen in the flavor basis rather than in the mass basis. In other words,
neutrinos propagate in their mass eigenstates but interact in their flavor eigenstate. The mixing of mass eigenstates during propagation
determines if the flavor eigenstate has oscillated or not.

The interaction mediated by the $W$ boson will give rise to an effective matter potential $V_{CC}$, while
the $Z$ boson is responsible for $V_{NC}$.

@@

            &= \sqrt{2}G_F N_e \bar{\nu}_{eR} \gamma^0 \nu_{eL}\,,

@@

where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand compontent of the neutrino fields interact.
Here we see a crucial difference that we expected: comparing the eigenvectors between the vacuum Hamiltonian defined in Eq.~\ref{eq:osc_1} and $H_{CC}$ shows us 
that by this construction, CC interactions occur in the flavor basis.

The effective 
potential that the electron neutrino experiences is then
\begin{align}\label{eq:V_CC1}
    V_{CC} = \sqrt{2}G_F N_e\,.
\end{align}

@@

Since only $\ne$ undergo CC interactions in Earth-like matter, the $V_{CC}$ potential is zero for all other flavors. 
However, since all flavors undergo NC interactions the total matter potential in the flavor basis is
\begin{align}\label{eq:V_matrix}
    V = \begin{bmatrix}
        V_{CC} + V_{NC} & 0 & 0 \\
        0 & V_{NC} & 0 \\
        0 & 0 & V_{NC} \\
    \end{bmatrix} = V_{CC}\, \delta_{\alpha e}\mathbb{I} + V_{NC}\mathbb{I}\,.
\end{align}
Just as in Eq.~\ref{eq:TDSE}, we start with a Hamiltonian that solves the time-dependent Schrödinger equation. This time, let the Hamiltonian be 
\begin{align}\label{eq:H_full}
    H = H_0 + H_{I}\,,
\end{align}
where $H_0$ is the Hamiltonian in vacuum, and $H_{I}$ is our interaction Hamiltonian associated with our matter potentials.
Let the amplitude for the $\nu_\a \to \nu_\b$ transition be denoted as
\begin{align}
    \psi_{\a\b}(t) = \bra{\nu_\b}\ket{\nu_\a (t)}\,,
\end{align}
i.e.~the evolution of the state of a neutrino emitted at $t =0$ with flavor $\alpha$ to flavor $\beta$ at time $t$. Now using Eq.~\ref{eq:osc_1} and Eq.~\ref{eq:V_CC1}, we are ready to see what form our Hamiltonians take.
Let us start with the vaccum Hamiltonian $H_0$, and act on its Schrödinger equation with $\bra{\nu_\b}\,$:
\begin{align}\label{eq:TDSE_H0}
    i \dv{}{t}\ket{\nu_\a (t)} = H_0\ket{\nu_\a (t)} \implies i \dv{}{t}\psi_{\a\b} = \bra{\nu_\b}H_0\ket{\nu_\a (t)}\,.
\end{align}
Eq.~\ref{eq:TDSE_H0} with the full Hamiltonian from Eq.~\ref{eq:H_full} is what we will numerically solve for the three neutrino picture
for neutrino oscillations in matter.

\subsection{Three Neutrino Picture}
Reminding ourselves that the vacuum Hamiltonian $H_0$ has eigenstates in the mass basis, we write the following expression where 
use the relations Eq.~\ref{eq:osc_1} and Eq.~\ref{eq:osc_k} to switch between the flavor and mass basis with the PMNS elements:
\begin{align}
    \bra{\nu_\b} H_0 &= \sum_k U_{\b k} \bra{\nu_k}H_0 \nonumber \\
                     &= \sum_k U_{\b k} E_k \bra{\nu_k} \nonumber \\
                     &= \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \bra{\nu_\eta}\,.
\end{align}
Thus,
\begin{align}
    \bra{\nu_\b}H_0\ket{\nu_\a (t)} &= \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \bra{\nu_\eta}\ket{\nu_\a (t)} \nonumber \\
                                    &= \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \psi_{\a\eta}(t)\,.
\end{align}
Using the ultrarelativistic approximation from Eq.~\ref{eq:ultra_rel}:
\begin{align}
    \sum_\eta \sum_k U_{\b k} E_k U^*_{\eta k} \psi_{\a\eta}(t) &= \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_k}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x) \nonumber \\
    &= \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_k}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x)\,.
\end{align}
Use the fact that $\sum_k m_k^2 =  \sum m_1^2 +_k m^2_k - m^2_1 =\sum_k m_1^2 +  \dm_{k1}$ to pull out common terms out of the summation:
\begin{align}\label{eq:t1}
    \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_k}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x) &= \sum_\eta \sum_k U_{\b k} \left(p + \frac{m^2_1}{2E} + \frac{\dm_{k1}}{2E}\right) U^*_{\eta k} \psi_{\a\eta}(x) \nonumber \\
    &= \sum_\eta\sum_k \left(p + \frac{m^2_1}{2E}\right) U_{\b k} U^*_{\eta k} \psi_{\a\eta}(x) \nonumber \\
    &+ \sum_\eta\sum_k U_{\b k} \frac{\dm_{k1}}{2E} U^*_{\eta k} \psi_{\a\eta}(x)\,.
\end{align}
Unitarity gives $ \sum_k U_{\b k} U^*_{\eta k} = \delta_{\eta\b}$, and the first term in the last step of Eq.~\ref{eq:t1} becomes
\begin{align}
    \sum_\eta \left(p + \frac{m^2_1}{2E}\right) \delta_{\beta \eta} \psi_{\a\eta}(x)
    =& \left(p + \frac{m^2_1}{2E}\right) \psi_{\a\b}(x)\,.
\end{align}

@@

\subsection{Two Neutrino Picture Solution}
By working in the limit where one of the neutrino masses can be neglected, 
we can reduce the complexity of the neutrino oscillations and obtain tractable analytical 
expressions. 
We now only consider the mixing between two neutrino flavors, say between $\ne, \nm$, and $\nu_1, \nu_2$. In other words, we let $\dm[31] = 0$.

The three-neutrino oscillation Hamiltonian from Eq.~\eqref{eq:H_3gen} %TODO: derive this
now reduces to 
\begin{align}
    H_F &= \frac{1}{2E}(U M^2 U^\dagger + A) \nonumber \\
        &= \frac{1}{2E}\left[U \begin{pmatrix}
            0 & 0 & 0 \\
            0 & \dm_{21} & 0 \\
            0 & 0 & 0
        \end{pmatrix} U^\dagger\right] + \sqrt{2}G_F N_e \begin{pmatrix}
            1 & 0 & 0 \\
            0 & 0 & 0 \\
            0 & 0 & 0
        \end{pmatrix}\,. 
\end{align}
The rotation matrices $R_{ij}$ that constitute $U$ according to Eq.~\eqref{PMNS_def} can be simplified. The rotation $R_{13}$ will not affect our mass matrix, since the former acts in the $1-3$ plane, and the latter sits in the $1-2$ plane. Thus, the reduced mass matrix $\mathbb{M}_{12}$ commutes with $R_{13}$. Analogously, the  rotation matrix $R_{23}$ also commutes with $\mathbb{M}_{12}$.
Moreover, since $SO(3)$ is abelian, its generators commute, i.e.
\begin{align}
    \commutator{R_{ij}}{R_{kl}} = R_{ij}R_{kl} - R_{kl}R_{ij} = 0\,.
\end{align}
Thus,
\begin{align}
    U \mathbb{M}_{12} U^\dagger &=
    R_{23}R_{13}R_{12} \mathbb{M}_{12}  R_{12}^\dagger R_{13}^\dagger  R_{23}^\dagger \nonumber \\
    &= R_{12}R_{23}R_{13} \mathbb{M}_{12}  R_{13}^\dagger R_{23}^\dagger R_{12}^\dagger   \nonumber \\
    &= R_{12}\mathbb{M}_{12} R_{23} R_{13}   R_{13}^\dagger R_{23}^\dagger R_{12}^\dagger  \,.
\end{align}
And due to the fact that the rotation matrices are unitary,
\begin{align}
    R\,R^\dagger = \mathbb{I}\,,
\end{align}
we have that 
\begin{align}
    U \mathbb{M}_{12} U^\dagger &= R_{12}\mathbb{M}_{12}R_{23} R_{13}   R_{13}^\dagger R_{12}^\dagger  R_{23}^\dagger \nonumber \\
    &= R_{12}\mathbb{M}_{12}R_{23} R_{23}^\dagger  R_{12}^\dagger \nonumber \\
    &= R_{12}\mathbb{M}_{12}  R_{12}^\dagger\,.
\end{align}
We have now completely removed the dependence of the mixing angle between the $\nu_1$ and $\nu_3$ mass eigenstates, $\theta_{13}$, and the dependece of the mixing angle between the $\nu_2$ and $\nu_3$ mass eigenstates, $\theta_{23}$.
We can see this clearly by writing our reduced mixing matrix, which is only constituted by the rotation in the $1-2$ plane, $R_{12}$.
\begin{align}
    U &= R_{12}
\begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1 \end{pmatrix} 
\end{align}

@@

Now we need to incorporate the \emph{zenith angle}, here defined as the angle between the neutrino direction of travel and south.
This way, neutrinos traveling through the entire diameter of the Earth are defined as `up-going', while
neutrinos that directly over-head are `down-going'. We will mostly work with the quantity $\cos{(\theta_z)}$.
Since we are interested in the matter effects, some studies are only looking at up-going neutrinos, i.e. neutrinos with
zenith angle $-1 \le \cos{(\theta_z)} \le 0$. However, while down-going neutrinos don't experience matter oscillations (and hardly even vacuum
oscillations due to the short baseline), they can be included to reduce systematic errors.
We now supplement our probability grid from Fig.~\ref{fig:oscillations} with the zenith dimension,
allowing us to fully see the Earth matter effect on the oscillations in full. This is shown in Fig.~\ref{fig:oscillograms}.

### 2021-11-16 thesis/tex/sterileTheory.tex
update

%\input{commands}
%\begin{document}
\section{The Sterile State}\label{sec:anomalies}
In 1996, the LSND experiment reported an excess of $\ane$ events from an $\anm$ beam~\cite{lsnd}.  
Nine years later, MiniBooNE not only reproduced the $\ane$ anomaly but observed an excess in the
$\ne$ events too. Together with the so-called reactor and gallium anomalies, these reports suggested 
that the three massive neutrino framework could be amended.
By introducing a fourth heavy mass state $\nu_4$, both appearance and disappearance anomalies could be minimally accommodated.
However, we know from the decay width of the $Z$ boson that it only can interact with three flavor species\footnote{
    Direct measurements of the invisible $Z$ decay yields $N=2.92 \pm 0.05$~\cite{pdg}.
}, 
so this fourth mass state cannot be interacting weakly. In other words, it needs to transform as a
singlet under the broken gauge group $SU(2)_L \times U(1)_{EM}$.
We now distinguish between the three original neutrino flavors ($e$, $\mu$, and $\tau$) and the new fourth 
flavor by calling the former \emph{active} neutrinos
and the latter \emph{sterile} due to its non-interacting behavior.

@@

\subsection{Sterile Signals}
Since the new neutrino does not interact weakly, how do we then detect its signal? If the sterile mixing angle $\theta_{i4}$ is non-zero, 
we allow the sterile mass state to mix with the active state $i$. The most interesting case is when $\theta_{24} \neq 0$, 
which for $\dm[41] \sim \si{\eV \squared}$ gives rise to a resonant disappeareance in the for $P_{\bar{\mu}\bar{\mu}}$, shown in Fig.~\ref{fig:sterile_resonance}.

>>>>

%\input{commands}
%\begin{document}
\section{The Sterile State}\label{sec:anomalies}
In 1996, the LSND experiment reported an excess of $\ane$ events from an $\anm$ beam~\cite{lsnd}.  
Nine years later, MiniBooNE not only reproduced the $\ane$ anomaly but observed an excess in the
$\ne$ events too. Together with the so-called reactor and gallium anomalies, these reports suggested 
that the three massive neutrino framework could be amended.
By introducing a fourth heavy mass state $\nu_4$, both appearance and disappearance anomalies could be minimally accommodated.
However, we know from the decay width of the $Z$ boson that it only can interact with three flavor species\footnote{
    Direct measurements of the invisible $Z$ decay yields $N=2.92 \pm 0.05$~\cite{pdg}.
}, 
so this fourth mass state cannot be interacting weakly. In other words, it needs to transform as a
singlet under the gauge symmetry group $SU(2)_L \times U(1)_{Y}$.
We now distinguish between the three original neutrino flavors ($e$, $\mu$, and $\tau$) and the new fourth 
flavor by calling the former \emph{active} neutrinos
and the latter \emph{sterile} due to its non-interacting behavior.

@@

\subsection{Effect of Sterile Neutrino on Probabilities}
Since the new, sterile, neutrino does not interact weakly, how do we know if it's there? If the sterile mixing angle $\theta_{i4}$ is non-zero, 
we allow the sterile mass state to mix with the active state $i$. The most interesting case is when $\theta_{24} \neq 0$, 
which for $\dm[41] \sim \si{\eV \squared}$ gives rise to a resonant disappeareance in the for $P_{\bar{\mu}\bar{\mu}}$, shown in Fig.~\ref{fig:sterile_resonance}.
